---
title: "Areas of endemism"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```

# Background
Areas of endemism are often of particular interest for biogeography and conservation. One qay to quantify endemsim is wieghted endemism (Laffan et al, 2001, https://doi.org/10.1046/j.1365-2699.2001.00524.x) as implemented in R (Guerin et al 2015a, https://doi.org/10.1111/2041-210X.12361) with option to correct forcorrection for total species richness.


# Tutorial
Usually more detailed distribution maps (for instsance from species distribution models) would be used, bu here we will use the cleaned data from www.gbif.org as an example. 


# Load libraries and some background data
```{r}
library(raster)
library(tidyverse)
library(viridis)
library(ggthemes)
library(cowplot)
library(rgdal)
#install.packages("input/adehabitat_1.8.20.tar.gz", repos = NULL, type = "source") 

wgs1984 <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
behr <- '+proj=cea +lon_0=0 +lat_ts=30 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +units=m +no_defs'

be <- raster("input/ABROCOMIDAE_ABROCOMIDAE.tif")
```

```{r}
## reproject
wgs1984 <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
behr <- '+proj=cea +lon_0=0 +lat_ts=30 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +units=m +no_defs'

be <- raster("input/ABROCOMIDAE_ABROCOMIDAE.tif")

# Weightened endemism following Guerin et al (https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12361)

## get occurrence recods from the ranges for each raster cell



# LONG ANALYSIS!
#bromeli <- readOGR(dsn = "output/bromeliaceae_ranges_with_data", layer = "bromeliaceae_ranges_with_data")
load("output/bromeliaceae_ranges_with_data.rda")

bromeli <- as(bromeli, "Spatial")
bromeli <- spTransform(bromeli, CRS(behr))

we_in <-  data.frame()

for(i in 1:nrow(bromeli)){
  print(i)
  ras <- bromeli[i,]
  ras <- rasterize(ras, be, getCover = TRUE)
  #num <- 1 / sum(getValues(ras) >= 0.25)
  ras[ras < 0.25] <- 0
  ras[ras > 0.25] <- 1

  out <- data.frame(
    species = bromeli[i,]$tax_accepted_name,
    rasterToPoints(ras))%>%
    filter(layer > 0 ) %>%
    dplyr::select(-layer)

  we_in <- bind_rows(we_in, out)
}

save(we_in, file = "output/occurrences_from_ranges_for_we.rda")


# convert to behrmann
load("output/occurrences_from_ranges_for_we.rda")
we <- weighted.endemism(species_records = we_in,
                        species="species", 
                        longitude="x",
                        latitude="y",
                        frame.raster = be)

plot(we$WE_raster)
writeRaster(we$WE_raster, 
            filename = "output/weightend_endemism_raster",
            overwrite = T, 
            format = "ascii")

# compare to null distribution

we_test <- endemism.null.test(we, nrep = 100, outlier.range = 1.5)

writeRaster(we_test$P.above.raster, 
            filename = "output/weightend_endemism_raster_pvalues",
            overwrite = T, 
            format = "ascii")


# BACkground maps
world.inp  <- rnaturalearth::ne_download(scale = 50,
                                         type = 'land',
                                         category = 'physical',
                                         load = TRUE)

count.inp  <- rnaturalearth::ne_download(scale = 110,
                                         type = 'countries',
                                         load = TRUE)

bra <- readOGR(dsn = "input", layer = "ne_10m_admin_1_states_provinces_lines")
bra <- bra[bra@data$adm0_a3 == "BRA",]
bra <-  raster::getData("GADM", country = "BRA", level = 1)

world.behr <- spTransform(world.inp, CRS(behr)) %>% fortify()
count.behr <- spTransform(count.inp, CRS(behr)) %>% fortify()
bra.behr <- spTransform(bra, CRS(behr)) %>% fortify()

# Supplementary figure weighted endemism
plo <- data.frame(rasterToPoints(we$WE_raster))%>%
  rename(we = ABROCOMIDAE_ABROCOMIDAE) %>% 
  filter(we > 0 )

plo_we <- ggplot()+
  geom_tile(data = plo, aes(x = x, y = y, fill = we))+
  scale_fill_viridis(name = "Weighted endemism", direction = -1, na.value = "transparent")+
  geom_polygon(data = world.behr,
               aes(x = long, y = lat, group = group), fill = "transparent", color = "black")+
  xlim(-12000000, -3000000)+
  ylim(-6500000, 4500000)+
  coord_fixed()+
  #ggtitle("Corrected Weighted Endemism")+
  theme_map()+
  theme(legend.justification = c(0, 0), 
        legend.position = c(0, 0))

ggsave(filename = "output/submission/Supplementary_figure_weighted_Endemism.jpg", plot = plo_we, height = 11, width = 8)

# Figure genus richness and endemsim test

ri_gen <-  raster("output/submission/richness_raster_bromeliaceae_genera_ranges")

plo_gen <- data.frame(rasterToPoints(ri_gen))%>%
  filter(layer > 0 )

plo_gen <- ggplot()+
  # geom_polygon(data = world.behr,
  #              aes(x = long, y = lat, group = group), fill = "transparent", color = "black")+
  geom_tile(data = plo_gen, aes(x = x, y = y, fill = layer), alpha = 1)+
  geom_polygon(data = count.behr,
               aes(x = long, y = lat, group = group), fill = "transparent", color = "grey20")+
  geom_polygon(data = bra.behr,
               aes(x = long, y = lat, group = group), fill = "transparent", color = "grey20")+
  scale_fill_viridis(name = "Genera", direction = -1, na.value = "transparent")+
  xlim(-12000000, -3000000)+
  ylim(-6500000, 4500000)+
  coord_fixed()+
  #ggtitle("Bromeliaceae")+
  theme_map()+
  theme(legend.justification = c(0, 0), 
        legend.position = c(0, 0))


plo_wep <- data.frame(rasterToPoints(we_test$P.above.raster))%>%
  rename(we_p = ABROCOMIDAE_ABROCOMIDAE) %>% 
  filter(we_p > 0 ) %>% 
  mutate(we_pcat = ifelse(we_p < 0.025, "High", we_p)) %>%
  mutate(we_pcat = ifelse(we_p > 0.975, "Low", we_pcat)) %>% 
  mutate(we_pcat = ifelse(we_p < 0.975 & we_p > 0.025, "Not significant", we_pcat))


plo_wep <- ggplot()+
  # geom_polygon(data = world.behr,
  #              aes(x = long, y = lat, group = group), fill = "transparent", color = "black")+
  geom_tile(data = plo_wep, aes(x = x, y = y, fill = we_pcat), alpha = 1)+
  geom_polygon(data = count.behr,
               aes(x = long, y = lat, group = group), fill = "transparent", color = "grey20")+
  geom_polygon(data = bra.behr,
               aes(x = long, y = lat, group = group), fill = "transparent", color = "grey20")+
  #scale_fill_viridis(name = "Genera", direction = -1, na.value = "transparent")+
  scale_fill_brewer(name = "Weighted Endemism", palette = "Set1")+
  xlim(-12000000, -3000000)+
  ylim(-6500000, 4500000)+
  coord_fixed()+
  #ggtitle("Bromeliaceae")+
  theme_map()+
  theme(legend.justification = c(0, 0), 
        legend.position = c(0, 0))


plo_end <- plot_grid(plo_gen, plo_wep,
                      labels = c('(a) Genus richness',
                                 '(b) Weighted endemism'), ncol = 2, hjust = 0)


ggsave(filename = "output/submission/Figure_genus_richness_weighted_endemism.pdf", plot = plo_end, dpi = 600)
```
