---
output: html_document
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```

# Objective
After this short exercise you will be able to download conservation assessments for multiple species from the INternational Union for the Conservation of Nature (www.iucn.org)

# Input data needed
- a taxon name for which to obtain fossils
- a list of species or genera names

# Exercises
1. Download existing conservation assessments from www.iucn.org for your group of interest using the IUCN API.

2. Familiarize yourself with the data 

# Tutorial

To obtain the the IUCN status of species using the r package [rredlist](https://cran.r-project.org/web/packages/rredlist/index.html), you need a token for the IUCN Redlist API.


```{r}
library(tidyverse)
library(readxl)
library(jsonlite)
library(rredlist)

iucn.key <- "YOUR TOKEN"

splist <- read_csv("example_data/day1_biodiversity_databases/bombacoideae_specieslist.csv") %>% 
  pull(species)

splist <- read_csv("example_data/day1_biodiversity_databases/species_list_sa_malvaceae.csv") %>% 
  rename(species = mal) %>%
  filter(!is.na(species)) %>% 
  pull(species)
   

iucn <- data.frame()

#get conservation status from IUCN via the IUCN API
for(i in 778:length(splist)){
  print(i)
  pick <- jsonlite::fromJSON(rl_search_(splist[i], key = iucn.key))$result

  # write.table(pick, "inst/secondary_woodyness_iucn_criteria.txt", 
  #             append = T, col.names = F, row.names = F)
  
  iucn <-  bind_rows(iucn, pick)
  Sys.sleep(1)
}

# check the data
head(iucn)
table(iucn$category)

# write to disk
write_csv(iucn, "bombacoideae_iucn_redlist_assessments.csv")
```

# Output generated
A .csv file with the conservation assessment for the species of your group of interest

# Further reading
https://cran.r-project.org/web/packages/rredlist/rredlist.pdf