---
title: "SPeceis ranges and richness"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval=FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```

# Exercises
1. Calculate the area of the EOO for all species in your group (`CalcRangeSize`)
2. Do a preliminary conservation assessment of your group based on Criterion B using t

# Possible questions for your project
* What is the maximum range size for a species in your group
* What does the distribution of range sizes look like? Is it normally distributed?

# Tutorial
## 1. Approximate species ranges

```{r}
dat <- read_csv("inst/occurrence_records_clean.csv")%>%
  dplyr::select(species, 
                decimallongitude = decimalLongitude,
                decimallatitude = decimalLatitude)

# Based on EOO
rs <- CalcRangeSize(dat)
```

A geospheric convex hull is a first approximation for a species range. However, some simple refinement might be desirable, for instance to limit the range only to biome where a given species has been recorded.

```{r}
# Limited to biomes with records
## Load Olson et al 2001 biomes
biom <- WWFload(x = "inst")
names(biom)

rs_biome <- CalcRangeSize(dat, biome = biom)

range <- data.frame(rs, rs_biome)
```
